---
title: "BLAST searches using orthologr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BLAST searches using orthologr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
library(orthologr)
has_blast <- nzchar(Sys.which("blastp"))
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  error = FALSE,
  tidy = FALSE,
  eval = has_blast)
```

* [1. Getting started with BLAST](#getting-started)
* [2. Core BLAST+ Functionality](#using-orthologr-to-perform-blast-searches)
    - [2.1 The blast() function](#the-blast-function)
        - [2.1.1 Inferring with one-to-one homologous hits](#working-with-one-to-one-hits)
        - [2.1.2 Inferring with one-to-many homologous hits](#working-with-one-to-many-hits)
    - [2.2 The blast_best() function](#the-blast_best-function)
    - [2.3 The blast_rec() function](#the-blast_rec-function)
    - [2.4 The set_blast() function](#the-set_blast-function)
* [3. Perform BLAST+ searches between genomes](#perform-blast-searches-between-genomes)
    - [3.1 Genome and Proteome Retrieval](#genome-and-proteome-retrieval)

## Getting Started

The `orthologr` package provides several interface functions to perform
BLAST searches.

__Note:__ If your goal is orthology inference or dN/dS estimation, consider using [DIAMOND2](https://github.com/bbuchfink/diamond) instead of BLAST — it is orders of magnitude faster and is the default aligner in `orthologr`. See the [Orthology Inference](https://drostlab.github.io/orthologr/articles/orthology_inference.html) and [dN/dS Estimation](https://drostlab.github.io/orthologr/articles/dNdS_estimation.html) vignettes for details. The functions documented here (`blast()`, `blast_best()`, `blast_rec()`) are retained for cases where BLAST output specifically is needed.

First, users need to make sure that they have [BLAST](ftp://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/) installed on their machine. Please follow [these](https://drostlab.github.io/orthologr/articles/Install.html#install-blast) instructions to install BLAST on your machine.

## Performing BLAST Searches

The `orthologr` package stores 20 example genes (orthologs) between _Arabidopsis thaliana_
and _Arabidopsis lyrata_. The following example BLAST search shall illustrate a simple
 search with standard parameters provided by the `blast()` function.

When running the subsequent functions please make sure you can call BLAST+ from your
console either in the standard `PATH` or in case you have BLAST+ installed in a separate
folder, please specify the `path` argument that can be passed to `blast()`.

To check whether BLAST+ can be executed from the default `PATH` (`usr/bin/local` on UNIX systems),
you can run:

```{r,eval=has_blast}
# check if blast is installed and if yes, what version of blast
system("blastp -version")
```

This should return something like this:

If everything works properly, you can get started with your first BLAST+ search.

## Using orthologr to perform BLAST searches

The `orthologr` packages allows users to perform fast and easy-to-use BLAST searches
between `fasta` files. These searches can even be scaled towards genome and proteome comparisons. 

### The blast() function

The `blast()` function provides the easiest way to perform a BLAST search.

```{r,eval=has_blast}
library(dplyr)
# performing a BLAST search using blastp (default)
hit_tbl <- blast(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                 subject_file = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'))
# look at results
glimpse(hit_tbl)
```

As you can see, the hit table shows the output of the BLAST+ search. The `blast()` function
runs `blastp` as default BLAST+ algorithm. Different BLAST+ algorithms can be selected by
specifying the `blast_algorithm` argument, e.g. `blast_algorithm = "tblastn"`. See `?blast` for further details.
The `blast()` function returns 21 columns: `query_id`, `subject_id`, `perc_identity`, `num_ident_matches`,
`alig_length`, `mismatches`, `gap_openings`, `n_gaps`, `pos_match`, `ppos`, `q_start`, `q_end`, `q_len`,
`qcov`, `qcovhsp`, `s_start`, `s_end`, `s_len`, `evalue`, `bit_score`, and `score_raw`.

Since `blast()` stores the hit table returned by BLAST in a data.table object, you can access each column,
using the [data.table notation](https://r-datatable.com).


In case you need to specify the `PATH` to BLAST+ please use the `path` argument:

```{r,eval=FALSE}
# performing a BLAST search using blastp (default)
hit_tbl <- blast(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                 subject_file = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
                 path         = "/path/to/blastp")


hit_tbl
```


```{r,eval=has_blast}
# access columns: query_id, subject_id, evalue, and bit_score
dplyr::select(hit_tbl, query_id, subject_id, evalue, bit_score)
```

The `blast()` function also allows you to pass additional parameters to the BLAST+ search
using the `blast_params` argument. In the following example, a remote BLAST+ search is
performed.

```r
hit_tbl <- blast(query_file = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                 subject_file = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
                 blast_params = "-qcov_hsp_perc 0.9")
```

```r
glimpse(hit_tbl)
```

```
Rows: 20
Columns: 21
$ query_id          <chr> "AT1G01010.1", "AT1G01020.1", "AT1G01030.1", "AT…
$ subject_id        <chr> "333554|PACid:16033839", "470181|PACid:16064328"…
$ perc_identity     <dbl> 73.987, 91.057, 95.543, 91.980, 100.000, 89.506,…
$ num_ident_matches <int> 347, 224, 343, 1812, 213, 580, 348, 271, 420, 49…
$ alig_length       <int> 469, 246, 359, 1970, 213, 648, 366, 300, 434, 52…
$ mismatches        <int> 80, 22, 12, 85, 0, 58, 14, 22, 8, 34, 4, 6, 68, …
$ gap_openings      <int> 8, 0, 2, 10, 0, 5, 2, 2, 3, 0, 0, 1, 3, 2, 1, 1,…
$ n_gaps            <int> 42, 0, 4, 73, 0, 10, 4, 7, 6, 0, 0, 1, 10, 8, 14…
$ pos_match         <int> 370, 231, 346, 1853, 213, 601, 351, 279, 425, 50…
$ ppos              <dbl> 78.89, 93.90, 96.38, 94.06, 100.00, 92.75, 95.90…
$ q_start           <int> 1, 1, 1, 6, 1, 1, 1, 1, 1, 1, 1, 1, 5, 4, 2, 1, …
$ q_end             <int> 430, 246, 359, 1910, 213, 646, 366, 294, 429, 52…
$ q_len             <int> 430, 246, 359, 1910, 213, 646, 366, 294, 429, 52…
$ qcov              <dbl> 100, 100, 100, 99, 100, 100, 100, 100, 100, 100,…
$ qcovhsp           <dbl> 100, 100, 100, 99, 100, 100, 100, 100, 100, 100,…
$ s_start           <int> 1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 16, 2, 4, 1,…
$ s_end             <int> 466, 246, 355, 1963, 213, 640, 362, 299, 433, 52…
$ s_len             <int> 466, 246, 355, 1963, 213, 640, 362, 299, 433, 52…
$ evalue            <dbl> 0.00e+00, 8.91e-168, 0.00e+00, 0.00e+00, 3.27e-1…
$ bit_score         <dbl> 627, 454, 698, 3704, 437, 1037, 696, 491, 859, 9…
$ score_raw         <dbl> 1617, 1169, 1801, 9604, 1125, 2681, 1797, 1264, …
```

In all cases the default `e-value` BLAST+ searches is `1E-5` and the default `blast_algorithm`
is `blastp`. 


Since BLAST+ searches can be computationally expensive, it is possible to specify the
`comp_cores` argument when working with an multicore machine.


```{r,eval=has_blast}

# BLAST computations using the comp_cores parameter: here with 2 cores
blast(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
      subject_file = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
      comp_cores   = 2)

```

The `query_file` and `subject_file` arguments specify the path to the corresponding 
fasta files storing the `CDS` files, `amino acid` files, or `genome` files of the query organism and subject organism
of interest. Make sure that when using `CDS`files, `amino acid` files, or `genome` files the 
corresponding argument `seq_type` must be adapted according to the input data format.

Use :

* `CDS` files -> `seq_type = "cds"`
* `amino acid` files -> `seq_type = "protein"`
* `genome` files -> `seq_type = "dna"`

The `format` argument specifies the input file format, e.g. "fasta" or "gbk".
The `blast_algorithm` argument specifies the BLAST program (algorithm) that shall be
used to perform BLAST searches, e.g. "blastp","blastn","tblastn",etc. Again, the `eval`
argument defines the default e-value that shall be chosen as best hit threshold.


### Using the split-apply-combine strategy for a BLAST hit table

All `blast` functions implemented in `orthologr` can easily be processed using the
[split-apply-combine strategy](https://www.jstatsoft.org/htaccess.php?volume=40&type=i&issue=01&filename=paper) to detect 
for example `one-to-one`, `one-to-many`, and `many-to-many` gene homology relationships.

Here a simple example:

```{r,eval=has_blast}
library(dplyr)
# perform a blastp search of 20 A. thaliana genes against 1000 A. lyrata genes
hit_tbl <- blast(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                 subject_file = system.file('seqs/ortho_lyra_cds_1000.fasta', package = 'orthologr'))

# determine 'one-to-many' and 'one-to-one' gene relationships
rel_hit_tbl <- summarize(group_by(hit_tbl, query_id), n_genes = n())
# look at results
rel_hit_tbl
```

Now you can sort genes into classes: `one-to-one` and `one-to-many`.

```{r,eval=has_blast}
# classify into 'one-to-one' relationships
one_to_one <- filter(rel_hit_tbl, n_genes == 1)
# classify into 'one-to-many' relationships
one_to_many <- filter(rel_hit_tbl, n_genes > 1)
```

```{r,eval=has_blast}
# look at one_to_one
one_to_one 
```

```{r,eval=has_blast}
# look at one_to_many
one_to_many
```

Now we can treat classes: `one_to_one` and `one_to_many` differently:

### Working with one-to-one hits

We can now retrieve the blast hit entries for the one-to-one hits
by joining the tables `one_to_one` and `hit_tbl`.

```{r,eval=has_blast}
# look at the evalue, perc_identity, and alig_length of one_to_one genes
oo_genes <- dplyr::inner_join(one_to_one, hit_tbl , by = "query_id")
oo_genes
```

This way, users will have all blast information of `one_to_one` hits available.

### A real world application: Homology Inference

Using the `oo_genes` dataset above, users can use this `one_to_one` hits strategy to
detect homologous genes between species.

For example, we could restrict `one_to_one` hits to fulfill certain (stringent)
criteria to identify as a homologous hit. Here we choose the following
parameter constellation to achieve this goal: `one_to_one` genes must have a minimum `alignment length of 300`, a p`erc_identity of > 80 percent` and an `e-value < 1E-5`. 

```{r,eval=has_blast}
# filter for 'homologous' hits
true_orthologs <- dplyr::filter(oo_genes, evalue < 1e-5, perc_identity > 80, alig_length > 300)
# look at results
dplyr::select(true_orthologs, query_id, subject_id, evalue, perc_identity, alig_length)
```

This way we could filter out a high confidence set of homologous genes from the
`one_to_one` class of genes.

In reality most homology inference programs and methods perform way more
complicated and sophisticated analyses to distinguish true orthologs from true paralogs
(in-paralogs, out-paralogs, etc.). These subsequent analyses can also be performed using
the above introduced split-apply-combine strategy. 

Note, that you can perform self-BLAST searches `blast(query,query)` and `blast(subject,subject)` to distinguish between orthologous and paralogous genes.

Now we continue with the `one_to_many` class of genes.

### Working with one-to-many hits

Here we want to address the question how to deal with multiple hits returned by `BLAST+`.

Again we investigate all `one_to_many` genes:

```{r,eval=has_blast}
# look at results 
one_to_many
```

When looking at gene_id `AT1G01200.1` we see that it was found 7 times in the
corresponding subject set of _A. lyrata_.

```{r,eval=has_blast}
# look at all 7 hits found
dplyr::select(dplyr::filter(hit_tbl, query_id == "AT1G01200.1"), query_id, subject_id, evalue, perc_identity, alig_length)
```

Now we have to decide which hit shall be considered as potential _homolog_.

In this example `subject_id` `470156|PACid:16041002` has the highest `perc_identity` as well
as the lowest e-value `1.87e-172`. So a straightforward approach would be to choose subject gene `470156|PACid:16041002` as potential ortholog of query gene `AT1G01200.1`.

We can validate this approach by running a reciprocal best hit search with `blast_rec()`and compare the output of gene `AT1G01200.1` with our choice `470156|PACid:16041002`.

A reciprocal best hit blast approach denotes the search strategy in which query sequences are blasted in one direction to detect matches in subject sequences and then inversely subject sequences are blasted in the other direction to detect matches in the query sequences. Only when both blast seaches in both directions result in the same hit pair: BLAST(A,B) = BLAST(B,A) 
the hit will be considered as true ortholog relationship.

```{r,eval=has_blast}
# run blast best reciprocal hit function
rbh_hit_tbl <- blast_rec(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                         subject_file = system.file('seqs/ortho_lyra_cds_1000.fasta', package = 'orthologr'))
# look at results
dplyr::select(rbh_hit_tbl, query_id, subject_id, evalue, perc_identity, alig_length)
```

When we now look at gene `AT1G01200.1` we find that after an reciprocal blast approach subject gene `470181|PACid:16064328` rather than the subject gene `470156|PACid:16041002` (= result of unidirectional blast) has been detected as potential ortholog. The example illustrates the importance of using bidirectional blast to determine orthology relationships. 

An alternative analysis that can be performed with these three candidate subject genes is the following:

```{r,eval=has_blast}
# read CDS sequences of the 20 example query genes of A. thaliana
Ath.cds <- orthologr::read.cds(file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'), format = "fasta")
# read CDS sequences of the 1000 example subject genes of A. lyrata
Aly.cds <- orthologr::read.cds(file   = system.file('seqs/ortho_lyra_cds_1000.fasta', package = 'orthologr'), format = "fasta")
# show the sequence of gene AT1G01070.1
Ath.cds["AT1G01070.1" , seqs]
```

Now users can perform a global alignment between the CDS sequences of `AT1G01070.1`
and the three subject genes as follows:

```{r,eval=has_blast}
library(Biostrings); library(pwalign)
# perform 3 global alignments between:  AT1G01070.1 and 918864|PACid:16052578, 
# 919693|PACid:16048878, 919961|PACid:16062329
sapply(Aly.cds[ unlist(dplyr::select(dplyr::filter(hit_tbl, query_id == "AT1G01070.1"), subject_id)), seqs ], pairwiseAlignment, 
       pattern = Ath.cds["AT1G01070.1" , seqs], type    = "global" )
```

__Note__: To obtain the score value, you need to specify the `scoreOnly = TRUE` in the `pairwiseAlignment` function.

As you can see, subject gene `918864|PACid:16052578` also has the highest global alignment score `1768.965` based on the Needleman-Wunsch algorithm. This strategy might help you to differentiate between border line cases.

The examples shown above shall demonstrate the use cases that can be performed using the
`blast` functions implemented in `orthologr`.


Another useful analysis can be to take the length of the initial query genes
into account using the `nchar()` function:

```{r,eval=has_blast}
# show the length distribution of all genes
# stored in "Ath.cds"
Ath.cds[ , nchar(seqs)]

```

Or the length of a specific gene:

```{r,eval=has_blast}

Ath.cds["AT1G01070.1" , nchar(seqs)]

```

This way you can easily visualize the length distribution of genes stored
in your query organism file.

```{r,eval=has_blast}
Ath.cds <- read.cds(system.file('seqs/ortho_thal_cds_1000.fasta', package = 'orthologr'),
                    format = "fasta")
                    
# look at sequence length distributions
hist(Ath.cds[ , nchar(seqs)], breaks = 100)
```

## The blast_best() function

For some analyses it is sufficient to perform BLAST+ best hit searches.
The `blast_best()` function is optimized to perform BLAST+ best hit searches
(only based on the minimum e-value) and returns the best hit when performing a BLAST+ search of a query organisms
(or set of query genes) against a subject organism (or set of subject genes).


```{r,eval=has_blast}
# performing gene orthology inference using the best hit (BH) method
blast_best(query_file    = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
           subject_file  = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
           clean_folders = TRUE)
```

The `blast_best()` function returns 21 columns: the same format as `blast()`. The most commonly used columns are `query_id`, `subject_id`, `perc_identity`, `alig_length`, `evalue`, and `bit_score`.

```{r,eval=has_blast}
# BLAST+ best hit search
best_hit_tbl <- blast_best(query_file      = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                           subject_file    = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'))
# look at results
dplyr::glimpse(best_hit_tbl)
```

## The blast_rec() function

The `blast_rec()` function was implemented to optimize BLAST+ reciprocal best hit searches
(only based on the minimum e-value).
BLAST+ reciprocal best hit searches are used to perform orthology inference.

Running `blast_rec()` using default parameter settings:

```{r,eval=has_blast}
# performing gene orthology inference using the reciprocal best hit (RBH) method
blast_rec(query_file   = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
          subject_file = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'))
```

## The set_blast() function

The `set_blast()`function reads a file storing a specific sequence type, such as "cds", "protein", or "dna" in a standard sequence file format such as "fasta", etc.
and depending of the makedb parameter either creates a blast-able database, or returns the corresponding protein sequences as data.table object for further BLAST searches.


```{r,eval=has_blast}
# using set_blast() to generate a blastable sequence database
head(set_blast(file = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'))[[1]] , 2)
```

## Advanced BLAST+ analyses

For large-scale applications of BLAST+ analysis within R, we recommend users to see the [metablastr](https://github.com/drostlab/metablastr) package
which aims to provide an easy-to-use BLAST search framework
for massive genome comparisons.


