---
title: "Protein Clustering with DIAMOND DeepClust"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Protein Clustering with DIAMOND DeepClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
library(orthologr)
has_diamond <- nzchar(Sys.which("diamond"))
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE,
  eval = has_diamond)
```

[DIAMOND DeepClust](https://github.com/bbuchfink/diamond) is a cascaded, ultra-fast method for clustering protein sequences at scale. It extends DIAMOND2 with a multi-pass clustering strategy that retains sensitivity at low sequence identity, making it practical for large proteome comparisons where reciprocal best hit methods become computationally expensive.

`orthologr` has three functions for working with `diamond deepclust`:

- `deepclust()` — cluster a set of protein sequences into homologous groups
- `deepclust_annotate()` — attach metadata to clusters
- `deepclust_realign()` — refine the alignment of sequences within each cluster

All three functions work with protein sequences only.

## `deepclust()`

`deepclust()` takes a fasta file of protein sequences and returns a data frame mapping each sequence to a cluster ID.

```{r,eval=TRUE}
library(orthologr)

clusters <- deepclust(
  input_file = c(
    system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
    system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr')
  )
)

clusters
```

## `deepclust_annotate()`

`deepclust_annotate()` takes the output of `deepclust()` and attaches functional annotations to each cluster.

```{r,eval=TRUE}
annotated <- deepclust_annotate(
  input_file = c(
    system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
    system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr')
  )
)

annotated
```


or use a named list to control the file_name column labels

```{r,eval=TRUE}
annotated <- deepclust_annotate(
  input_file = list(
    "thal" = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
    "lyra" = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr')
  )
)
annotated
```

```{r,eval=TRUE}
# Stricter clustering with custom thresholds
annotated <- deepclust_annotate(
  input_file = c(
    system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
    system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr')
  ),
  approx_id    = 50,
  mutual_cover = 90,
  comp_cores   = 4
)
annotated
```

## `deepclust_realign()`

`deepclust_realign()` refines the multiple sequence alignment within each cluster, which can improve downstream dN/dS estimation or phylogenetic inference.

```{r,eval=TRUE}
realign_result <- deepclust_realign(
  input_file = c(
    system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
    system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr')
  ),
  clusters = clusters
)

realign_result
```

This feature is still under development. More information about the DeepClust algorithm can be found in the [DIAMOND DeepClust paper](https://www.nature.com/articles/s41592-026-03030-z).
