---
title: "Orthology inference dNdS estimation of orthologous genes using orthologr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Orthology inference dNdS estimation of orthologous genes using orthologr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
has_diamond <- nzchar(Sys.which("diamond"))
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE,
  eval = has_diamond)
```

The dN/dS ratio quantifies the mode and strength of selection acting on a pair of orthologous genes.
This selection pressure can be quantified by comparing synonymous substitution rates (dS) that are assumed to be neutral
with nonsynonymous substitution rates (dN), which are exposed to selection as they 
change the amino acid composition of a protein ([Mugal et al., 2013](https://academic.oup.com/mbe/article/31/1/212/1049986)).

The `orthologr` package provides a function named `dNdS()` to perform dNdS estimation on pairs of orthologous genes.
The `dNdS()` function takes the CDS files of two organisms of interest (`query_file` and `subject_file`) 
and computes the dNdS estimation values for orthologous gene pairs between these organisms.

__Note:__ the following dNdS estimation methods are based on KaKs_Calculator 2:

* "NG": Nei, M. and Gojobori, T. (1986)

* "LWL": Li, W.H., et al. (1985)

* "LPB": Li, W.H. (1993) and Pamilo, P. and Bianchi, N.O. (1993)

* "MLWL" (Modified LWL), MLPB (Modified LPB): Tzeng, Y.H., et al. (2004)

* "YN": Yang, Z. and Nielsen, R. (2000)

* "MYN" (Modified YN): Zhang, Z., et al. (2006)

* "GMYN": Wang, D.P., et al. Biology Direct. (2009)

* "GY": Goldman, N. and Yang, Z. (1994)

* "MS": (Model Selection): based on a set of candidate models, Posada, D. (2003) 
* "MA" (Model Averaging): based on a set of candidate models, Posada, D. (2003)

* "ALL": All models toghether 

It is assumed that when you choose one of these dNdS estimation methods you have
[KaKs_Calculator 2](https://sourceforge.net/projects/kakscalculator2/) installed on your machine and it can be executed from the
default execution `PATH`.

The following pipeline resembles an example dNdS estimation procedure:

 1) Orthology Inference: e.g. DIAMOND2 reciprocal best hit (RBH) — default; or BLAST reciprocal best hit
 
 2) Pairwise sequence alignment: e.g. clustalw for pairwise amino acid sequence alignments
 
 3) Codon Alignment: e.g. pal2nal program
 
 4) dNdS estimation: e.g. [Yang, Z. and Nielsen, R. (2000)](https://academic.oup.com/mbe/article/17/1/32/975527) (YN)


__Note:__ it is assumed that when using `dNdS()` all corresponding programs you
want to use are already installed on your machine and are executable via either
the default execution PATH or you specifically define the location of the executable program
via the `aa_aln_path` or `aligner_path` argument that can be passed to `dNdS()`. By default `dNdS()` uses DIAMOND2 (`aligner = "diamond"`); you can switch to BLAST by setting `aligner = "blast"`. See the [Sequence Alignments vignette](https://drostlab.github.io/orthologr/articles/sequence_alignments.html)
for details.

The following example shall illustrate a dNdS estimation process.


```{r,eval=TRUE}
library(orthologr)
# get a dNdS table using:
# 1) DIAMOND2 reciprocal best hit for orthology inference (default; aligner = "diamond", ortho_detection = "RBH")
# 2) Needleman-Wunsch for pairwise amino acid alignments
# 3) pal2nal for codon alignments
# 4) Comeron for dNdS estimation
# 5) single core processing 'comp_cores = 1'
dNdS(query_file      = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
     subject_file    = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
     aligner         = "diamond",
     ortho_detection = "RBH", 
     aa_aln_type     = "pairwise",
     aa_aln_tool     = "NW", 
     codon_aln_tool  = "pal2nal", 
     dnds_est.method = "Comeron", 
     comp_cores      = 1)
```
Some outputs include `NA` values. To filter for `NA` values or a specific `dnds.threshold`,
you can use the `filter_dNdS()` function. The `filter_dNdS()` function takes the output data.table 
returned by `dNdS()` and filters the output by the following criteria:

1) all dN values having an NA value are omitted

2) all dS values having an NA value are omitted

3) all dNdS values >= the specified `dnds.threshold` are omitted


```{r,eval=TRUE}
library(orthologr)
# get dNdS estimated for orthologous genes between A. thaliana and A. lyrata
Ath_Aly_dnds <- dNdS(query_file      = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
     subject_file    = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
     aligner         = "diamond",
     ortho_detection = "RBH", 
     aa_aln_type     = "pairwise",
     aa_aln_tool     = "NW", 
     codon_aln_tool  = "pal2nal", 
     dnds_est.method = "Comeron", 
     comp_cores      = 1)
# filter for:
# 1) all dN values having an NA value are omitted
# 2) all dS values having an NA value are omitted
# 3) all dNdS values >= 2 are omitted
filter_dNdS(Ath_Aly_dnds, dnds.threshold = 2)
```

The `dNdS()` function can be used choosing the following options:

* `aligner` : `"diamond"` (default; DIAMOND2) or `"blast"` (BLASTP)
* `ortho_detection` : `RBH` (reciprocal best hit), `BH` (best hit) — works with both DIAMOND2 and BLAST depending on `aligner`
* `aa_aln_type` : `multiple` or `pairwise`
* `aa_aln_tool` : `clustalw`, `t_coffee`, `muscle`, `clustalo`, `mafft`, and `NW` (in case `aa_aln_type = "pairwise"`)
* `codon_aln_tool` : `pal2nal`
* `dnds_est.method` : `Li`, `Comeron`, `NG`, `LWL`, `LPB`, `MLWL`, `YN`, and `MYN`

Please see `?dNdS` for details.

In case your DIAMOND2 program or multiple alignment program cannot be executed from the default execution `PATH`
you can specify the `aa_aln_path` or `aligner_path` arguments.

```{r,eval=FALSE}
library(orthologr)
# using the `aa_aln_path` or `aligner_path` arguments
dNdS( query_file      = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
      subject_file    = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
      aligner         = "diamond",
      ortho_detection = "RBH",
      aligner_path    = "here/path/to/diamond",
      aa_aln_type     = "multiple", 
      aa_aln_tool     = "clustalw", 
      aa_aln_path     = "here/path/to/clustalw",
      codon_aln_tool  = "pal2nal", 
      dnds_est.method = "Comeron", 
      comp_cores      = 1, 
      clean_folders   = TRUE)
```


## Advanced options

Additional arguments can be passed to `dNdS()`. This allows you to use more advanced 
options of several interface programs.

To pass additional parameters to the interface programs, you can use the
`blast_params` and `aa_aln_params` arguments. The `aa_aln_params` argument assumes
that when you chose e.g. `aa_aln_tool = "mafft"` you will pass the corresponding
additional parameters in MAFFT notation.

```{r,eval=FALSE}
library(orthologr)
# get dNdS estimated for orthologous genes between A. thaliana and A. lyrata
# using additional parameters:

# get a dNdS table using:
# 1) DIAMOND2 reciprocal best hit for orthology inference (RBH), with extra DIAMOND2 params
# 2) multiple amino acid alignments using MAFFT
# 3) pal2nal for codon alignments
# 4) Comeron (1995) for dNdS estimation
# 5) single core processing 'comp_cores = 1'
Ath_Aly_dnds <- dNdS( query_file      = system.file('seqs/ortho_thal_cds.fasta', package = 'orthologr'),
                      subject_file    = system.file('seqs/ortho_lyra_cds.fasta', package = 'orthologr'),
                      aligner         = "diamond",
                      ortho_detection = "RBH",
                      aligner_params  = "--matrix BLOSUM80",
                      aa_aln_type     = "multiple",
                      aa_aln_tool     = "mafft",
                      aa_aln_params   = "--maxiterate 1 --clustalout",
                      dnds_est.method = "Comeron",
                      comp_cores      = 1, 
                      clean_folders   = TRUE, 
                      quiet           = TRUE )

# filter for:
# 1) all dN values having an NA value are omitted
# 2) all dS values having an NA value are omitted
# 3) all dNdS values >= 0.1 are omitted
filter_dNdS(Ath_Aly_dnds, dnds.threshold = 0.1)
```


```

     query_id      subject_id       dN     dS    dNdS
1 AT1G01050.1 909874|PACid:16 0.000000 0.1750 0.00000
2 AT1G01090.1 470171|PACid:16 0.009843 0.2150 0.04579
3 AT1G01120.1 918858|PACid:16 0.003072 0.1326 0.02317
4 AT1G01140.3 470161|PACid:16 0.005672 0.1312 0.04324
5 AT1G01170.2 311317|PACid:16 0.008750 0.2827 0.03095
6 AT1G01220.1 470155|PACid:16 0.015210 0.1533 0.09919

```
Here `aligner_params` and `aa_aln_params` take a character string specifying the parameters
that shall be passed to DIAMOND2 (or BLAST when `aligner = "blast"`) and MAFFT, respectively.
The notation of these parameters must follow the command line call of the stand alone versions
of DIAMOND2 and MAFFT: e.g. `aligner_params = "--matrix BLOSUM80"`
and `aa_aln_params = "--maxiterate 1 --clustalout"`.

