---
title: "Orthology Inference using orthologr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Orthology Inference using orthologr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
library(orthologr)
has_diamond <- nzchar(Sys.which("diamond"))
has_blast   <- nzchar(Sys.which("blastp"))
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE,
  eval = has_diamond)
```

Orthology Inference is the process of detecting [orthologous genes](https://www.biostars.org/p/1595/) between a query organism and
a set of subject organisms. Motivated by the discussion on which orthology inference method, paradigm or program is 
the [most accurate](https://www.biostars.org/p/7568/) or [the fastest](https://www.nature.com/articles/nmeth.3830) one, the `orthologr` package provides functions to perform genome wide orthology inference following two different paradigms:

>1) orthology inference based on 1-1 orthology relationship using `DIAMOND2 reciprocal best hit` (default, recommended) or `BLAST reciprocal best hit`
>
>2) orthology inference based on 1-many homology relationship using orthologous groups (OrthoFinder2)


To perform orthology inference you can start with the `orthologs()` function provided by `orthologr`.
The `orthologs()` function takes nucleotide or protein sequences stored in fasta files for a set of organisms as input
and performs orthology inference to detect orthologous genes or orthologous groups within the given organisms based on selected orthology inference programs.

The following interfaces are implemented in the `orthologs()` function:

### DIAMOND2 based methods (recommended):

- DIAMOND2 reciprocal best hit (DIAMOND_RBH): see `diamond_rec()`. **This is the default** (`ortho_detection = "DIAMOND_RBH"`). DIAMOND2 is up to 10,000× faster than BLAST in default mode and retains full BLAST-level sensitivity in `ultra-sensitive` mode.

- DIAMOND2 best hit (DIAMOND_BH): see `diamond_best()`.

### BLAST based methods:

- BLASTp best hit (BH): see `blast_best()`.

- BLASTp reciprocal best hit (RBH): see `blast_rec()`.

### Orthogroup based methods:

- [Orthofinder2](https://github.com/davidemms/OrthoFinder): see `orthofinder2()`.


### Examples: DIAMOND2 based methods (default)

Using a simple example stored in the package environment of `orthologr` you
can get an impression on how to use the `orthologs()` function.

__Note:__ it is assumed that when using `orthologs()` all corresponding programs you
want to use are already installed on your machine and are executable via either
the default execution PATH or you specifically define the location of the executable file
via the `path` argument that can be passed to `orthologs()`. For DIAMOND2-based methods
(the default), make sure [DIAMOND2](https://github.com/bbuchfink/diamond) is installed.
See the [Installation Vignette](https://drostlab.github.io/orthologr/articles/Install.html) for details.

In the following examples, we will use toy fasta files stored within the `orthologr` package: `system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr')`.


```{r,eval=TRUE}
# perform orthology inference using DIAMOND2 reciprocal best hit (default)
# and fasta sequence files storing protein sequences
orthologr::orthologs(query_file    = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_RBH",
          comp_cores      = 1,
          clean_folders   = FALSE)
```


This small example returns 20 orthologous genes between _Arabidopsis thaliana_ and _Arabidopsis lyrata_.
As you can see, the `query_file` and `subject_files` arguments take the proteomes of _Arabidopsis thaliana_ (`query_file`) and _Arabidopsis lyrata_ (`subject_files`) stored in fasta files. The `seq_type` argument specifies that you will pass protein sequences (proteomes)
to the `orthologs()` function. In case you only have either genomes (DNA sequences) or CDS files, you can
modify the `seq_type` argument to `seq_type = "dna"` (when working with only genome data) or `seq_type = "cds"` (when working with CDS files).
Internally the `orthologs()` function will perform a CDS prediction using `predict_cds()` and will furthermore translate the predicted
CDS sequences into protein sequences. Analogously when `seq_type = "cds"` is specified, internally the `orthologs()` function will
translate all CDS sequences into protein sequences to run orthology inference based on protein sequences.

__The advantage of this type of output is that in addition to having the orthology relationship between genes from two different genomes, users also retrieve the DIAMOND2 alignment results of the respective orthology relationship (in the same format as BLAST) which allows
them to perform subsequent filtering for either more conservative or more liberal orthology relationships.__

__Note__: future versions of `orthologr` will allow to perform orthology inference using DNA sequences.
Nevertheless, since most orthology inference methods or paradigms rely on protein sequences, the first version
of `orthologr` will follow this paradigm.

In case you have to specify the path to DIAMOND2 you can use the `path` argument as follows:


```{r,eval=FALSE}
# using an external execution path for DIAMOND2
orthologr::orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_RBH",
          path            = "here/path/to/diamond",
          clean_folders   = FALSE,
          comp_cores      = 1)
```


When you are working on a multicore machine, you can also specify the `comp_cores`
argument that will allow you to run all analyses in parallel (to speed up computations).


```{r,eval=TRUE}
# running DIAMOND2 orthology inference in parallel using 2 cores
orthologr::orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_RBH",
          clean_folders   = FALSE,
          comp_cores      = 2)
```

In this case 2 cores are being used to perform parallel processing, the `clean_folders` argument
specifies that all files returned by the corresponding orthology inference method
are removed after analyses.

## Program specific use of the orthologs() function

In this section small examples will illustrate the use
of the `orthologs()` function for each orthology inference program.

### DIAMOND2 best hit

The DIAMOND2 best hit method is a uni-directional DIAMOND2 best hit search of a `query organism A`
against a `subject organism B` based on the `e-value`. It uses the same hit-filtering logic as `blast_best()` but is substantially faster.

Orthology Inference using DIAMOND2 best hit can be performed by specifying the argument `ortho_detection = "DIAMOND_BH"`
and one computing core `comp_core = 1`:

```{r,eval=TRUE}
# perform orthology inference using DIAMOND2 best hit
# and fasta sequence files storing protein sequences
orthologr::orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_BH",
          clean_folders   = TRUE,
          comp_cores      = 1)
```

The resulting table stores the orthologous gene pairs and the corresponding alignment statistics.
DIAMOND2 returns additional columns compared to BLAST (e.g. `n_gaps`, `pos_match`, `ppos`, `q_len`, `qcov`, `s_len`, `score_raw`).


### DIAMOND2 reciprocal best hit (default, recommended)

The DIAMOND2 reciprocal best hit is a bi-directional DIAMOND2 best hit search of a `query organism A`
against a `subject organism B` based on the `e-value`. This is the **default** method in `orthologs()`.

The algorithm runs as follows:

1) `bh_A <- diamond_best(A, B)`

2) `bh_B <- diamond_best(B, A)`

3) `join(bh_A, bh_B)` by tuple `(query_id, subject_id)`

Only when the tuple `(query_id, subject_id)` is returned as best hit in both DIAMOND2 directions
is it retained as an orthologous gene pair.

```{r,eval=TRUE}
library(orthologr)

# perform orthology inference using DIAMOND2 reciprocal best hit (default)
# and fasta sequence files storing protein sequences
orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_RBH",
          clean_folders   = FALSE,
          comp_cores      = 1)
```

You can also adjust the DIAMOND2 sensitivity level using the `sensitivity_mode` argument. The default
is `"fast"`, but for more distant species comparisons you may want `"sensitive"` or `"ultra-sensitive"`:

```{r,eval=TRUE}
library(orthologr)

# DIAMOND2 RBH with ultra-sensitive mode (equivalent to BLAST sensitivity)
orthologs(query_file       = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files    = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type         = "protein",
          ortho_detection  = "DIAMOND_RBH",
          sensitivity_mode = "ultra-sensitive",
          clean_folders    = FALSE,
          comp_cores       = 1)
```

The full set of sensitivity modes available in DIAMOND2 is:

| `sensitivity_mode` | Speed | Use case |
|---|---|---|
| `"faster"` | Fastest | Hits >70% identity |
| `"fast"` | Fast (default) | Hits >70% identity |
| `"mid-sensitive"` | Moderate | Between fast and sensitive |
| `"sensitive"` | Sensitive | Full sensitivity for hits >40% identity |
| `"more-sensitive"` | More sensitive | — |
| `"very-sensitive"` | Very sensitive | — |
| `"ultra-sensitive"` | BLAST-equivalent | Most sensitive; slowest |

To retrieve the full DIAMOND2 hit table for subsequent analyses, set `clean_folders = FALSE`.
The corresponding hit table can then be found in `file.path(tempdir(), "_blast_db")`.

```{r,eval=TRUE}
library(orthologr)
library(dplyr)

# perform orthology inference using DIAMOND2 reciprocal best hit
RBH <- orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein",
          ortho_detection = "DIAMOND_RBH",
          clean_folders   = FALSE,
          comp_cores      = 1)

glimpse(RBH)
```

### BLASTp best hit

The BLASTp best hit method is a uni-directional BLAST best hit search of a `query organism A` 
against a `subject organism B` based on the `e-value`.

Orthology Inference using BLASTp best hit can be performed by specifying the argument `ortho_detection = "BH"`
and one computing core `comp_core = 1`:

```{r,eval=has_blast}
# perform orthology inference using BLAST best hit
# and fasta sequence files storing protein sequences
orthologr::orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein", 
          ortho_detection = "BH", 
          clean_folders   = TRUE,
          comp_cores      = 1)
```

The resulting table stores the orthologous gene pairs and the corresponding BLAST alignment statistics.


### BLASTp best reciprocal hit

The BLAST best reciprocal hit is a bi-directional BLAST best hit search of a `query organism A` 
against a `subject organism B` based on the `e-value`.

The Algorithm for BLAST best reciprocal hit runs as follows:

1) `bh_A <- best_hit(A,B)`

2) `bh_B <- best_hit(B,A)`

3) `join(bh_A,bh_B)` by tupel `(query_id, subject_id)`

In other words, only in case the tuple `(query_id, subject_id)` is returned as best hit
based on the `e-value` in both BLAST directions, the corresponding tupel `(query_id, subject_id)`
is retained as orthologous gene pair.

```{r,eval=has_blast}

library(orthologr)

# perform orthology inference using BLAST reciprocal best hit
# and fasta sequence files storing protein sequences
orthologs(query_file      = system.file('seqs/ortho_thal_aa.fasta', package = 'orthologr'),
          subject_files   = system.file('seqs/ortho_lyra_aa.fasta', package = 'orthologr'),
          seq_type        = "protein", 
          ortho_detection = "RBH", 
          clean_folders   = FALSE,
          comp_cores      = 1)
```

In case you would like to store the corresponding `hit tables` returned by BLAST for
subsequent analyses, you can specify the `clean_folders = FALSE` argument. The corresponding
BLAST hit table can then be found in `file.path(tempdir(),"_blast_db")`.

A detailed overview of further analyses that can be done with the corresponding BLAST output
can be found in the [BLAST vignette](https://drostlab.github.io/orthologr/articles/blast.html).

<!-- ## Multi-species ortholog table workflow -->

<!-- The pairwise functions above compare one query against one subject at a time. When you want to -->
<!-- compare a query species against **several** subject species and then identify a **core set** of -->
<!-- orthologs conserved across all of them, `orthologr` provides a four-step pipeline: -->

<!-- | Step | Function | What it does | -->
<!-- |------|----------|--------------| -->
<!-- | 1 | `map_generator_dnds()` | Pairwise BLAST + dNdS between query and every subject; writes one CSV per species pair | -->
<!-- | 2 | `generate_ortholog_tables_all()` | Combines those CSVs with GTF/GFF annotation files to resolve gene loci from splice variants | -->
<!-- | 3 | `import_ortholog_tables_all()` | Re-reads saved ortholog tables from disk (skip if you kept the in-memory result from step 2) | -->
<!-- | 4 | `retrieve_core_orthologs()` | Retains only gene loci with an ortholog in **every** subject species | -->

<!-- ### Step 1: Generate pairwise dNdS maps -->

<!-- `map_generator_dnds()` iterates over all CDS files in `subjects_folder`, runs pairwise orthology -->
<!-- inference and dNdS estimation against the query, and writes one CSV per species pair to -->
<!-- `output_folder`. -->

<!-- ```{r, eval=FALSE} -->
<!-- library(orthologr) -->

<!-- map_generator_dnds( -->
<!--   query_file      = "path/to/query_cds.fasta", -->
<!--   subjects_folder = "path/to/subject_cds_files/", -->
<!--   output_folder   = "dnds_maps/", -->
<!--   ortho_detection = "DIAMOND_RBH", -->
<!--   aa_aln_type     = "pairwise", -->
<!--   aa_aln_tool     = "NW", -->
<!--   codon_aln_tool  = "pal2nal", -->
<!--   dnds_est_method = "Comeron", -->
<!--   comp_cores      = 1 -->
<!-- ) -->
<!-- ``` -->

<!-- The toy files bundled with the package can be used to test this: -->

<!-- ```{r, eval=FALSE} -->
<!-- map_generator_dnds( -->
<!--   query_file      = system.file("seqs/ortho_thal_cds.fasta", package = "orthologr"), -->
<!--   subjects_folder = system.file("seqs/map_gen_example",      package = "orthologr"), -->
<!--   output_folder   = "orthologr_dnds_maps/", -->
<!--   aa_aln_type     = "pairwise", -->
<!--   aa_aln_tool     = "NW", -->
<!--   codon_aln_tool  = "pal2nal", -->
<!--   dnds_est_method = "Comeron", -->
<!--   comp_cores      = 1 -->
<!-- ) -->
<!-- ``` -->

<!-- ### Step 2: Build gene-locus ortholog tables -->

<!-- `generate_ortholog_tables_all()` takes the dNdS CSVs from step 1 and a set of GTF/GFF annotation -->
<!-- files to resolve splice variants to gene loci, producing a combined ortholog table with columns -->
<!-- `query_species`, `subject_species`, `query_gene_locus_id`, `query_id`, `subject_id`, `dN`, `dS`, -->
<!-- `dNdS`, and standard alignment statistics. -->

<!-- `output_type = "gene_locus"` (default) retains one representative splice variant per gene locus — -->
<!-- the one with the lowest e-value (longest isoform used as tiebreaker). Use -->
<!-- `output_type = "splice_variant"` to keep all splice variant pairings. -->

<!-- ```{r, eval=FALSE} -->
<!-- ortho_tbl <- generate_ortholog_tables_all( -->
<!--   dNdS_folder               = "dnds_maps/", -->
<!--   annotation_file_query     = "path/to/query_annotation.gtf", -->
<!--   annotation_folder_subject = "path/to/subject_annotations/", -->
<!--   output_folder             = "ortholog_tables/", -->
<!--   output_type               = "gene_locus",    # or "splice_variant" -->
<!--   format                    = c("gtf", "gtf")  # adjust to "gff" as needed -->
<!-- ) -->
<!-- ``` -->

<!-- ### Step 3: Re-import saved tables (optional) -->

<!-- If you need to reload the ortholog tables in a later session without rerunning steps 1–2, use -->
<!-- `import_ortholog_tables_all()`: -->

<!-- ```{r, eval=FALSE} -->
<!-- ortho_tbl <- import_ortholog_tables_all(ortholog_tables_folder = "ortholog_tables/") -->
<!-- ``` -->

<!-- ### Step 4: Retrieve the core ortholog set -->

<!-- `retrieve_core_orthologs()` filters the combined table to gene loci that have an ortholog in -->
<!-- **every** subject species. Pass the subject species in phylogenetic order (closest relative first) -->
<!-- via `species_order`. -->

<!-- ```{r, eval=FALSE} -->
<!-- core_orthologs <- retrieve_core_orthologs( -->
<!--   ortho_tables = ortho_tbl, -->
<!--   species_order = c("Arabidopsis_lyrata", "Brassica_rapa", "Capsella_rubella") -->
<!-- ) -->
<!-- ``` -->

<!-- The returned table has the same schema as `ortho_tbl` but contains only gene loci conserved across -->
<!-- all species, making it a convenient starting point for comparative analyses such as divergence -->
<!-- stratigraphy (see the [Divergence Stratigraphy vignette](https://drostlab.github.io/orthologr/articles/divergence_stratigraphy.html)). -->
