---
title: "Setting up DIAMOND2 and BLAST+ on Windows"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up DIAMOND2 and BLAST+ on Windows}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  error = FALSE,
  tidy = FALSE)
```

* [1. Getting started](#getting-started)
* [2. Install Rtools](#install-rtools)
* [3. Confirm the location of the executables](#confirm-the-location-of-the-executables)
* [4. Set PATH for the current R session (temporary)](#set-path-for-the-current-r-session-temporary)
* [5. Verify the installation](#verify-the-installation)
* [6. Additional information](#additional-information)

## Getting Started

`DIAMOND2` and `BLAST+` are external executables that `orthologr` shells out
to (e.g. `diamond()`, `deepclust()`, `deepclust_realign()`, `blast()`,
`blast_best()`, `blast_rec()`). Neither is an R package, so on Windows both
must be installed separately and made discoverable to R before these
interface functions can be used. See the [Install](Install.html) vignette
for general installation instructions across platforms.

This vignette focuses on Windows, and describes the quickest way to make
both tools discoverable for a single R session without touching your
system-wide configuration: temporarily adding their folders to `PATH`.

## Install Rtools

This section is about building `orthologr` itself, not about DIAMOND2 or
BLAST+. `orthologr` ships compiled C++ code (used, for example, by
`gestimator()`), so installing it from source on Windows — via
`install.packages()`, `devtools::install_github()`, or
`devtools::load_all()` — _may_ require a working compiler toolchain. Windows does
not include one by default, so you need to install
[**Rtools**](https://cran.r-project.org/bin/windows/Rtools/).

1. Check your R version (`R.version.string`) and download the matching
   Rtools installer from the CRAN Rtools page (e.g. Rtools45 for R 4.5.x).
2. Run the installer with default settings.
3. Restart RStudio.
4. Verify the toolchain is available:

```{r,eval=FALSE}
pkgbuild::check_build_tools(debug = TRUE)
Sys.which("make")
```

If `make` resolves to an empty string, or `pkgbuild::has_build_tools()`
returns `FALSE`, Rtools was not installed correctly (a common cause is an
interrupted or incomplete installer run, which can leave an empty
`c:\rtoolsNN` folder despite `PATH` being set correctly) — reinstall Rtools
before continuing.

Once Rtools is confirmed working, `devtools::load_all()` and
`R CMD INSTALL` should be able to compile `orthologr`'s C++ sources.

## Confirm the location of the executables

Before doing anything else, confirm that `diamond.exe` and `blastp.exe`
actually exist where you expect them. Example locations:

```
C:/Users/<you>/Desktop/R_projects/installations/diamond.exe
C:/Program Files/NCBI/blast-2.17.0+/bin/blastp.exe
```

```{r,eval=FALSE}
file.exists("C:/Users/<you>/Desktop/R_projects/installations/diamond.exe")
file.exists("C:/Program Files/NCBI/blast-2.17.0+/bin/blastp.exe")
```

If `DIAMOND2` or `BLAST+` is not yet installed, see the
[Install](Install.html#install-blast) vignette for download links.

## Set PATH for the current R session (temporary)

This option adds each tool's folder to the `PATH` environment variable for
the duration of the current R session only. It's useful for one-off scripts 
without changing any system settings, and it avoids having to pass a 
`path` argument to every `orthologr` function call.

```{r,eval=FALSE}
Sys.setenv(
  PATH = paste(
    "C:/Users/<you>/Desktop/R_projects/installations",
    "C:/Program Files/NCBI/blast-2.17.0+/bin",
    Sys.getenv("PATH"),
    sep = ";"
  )
)
```

This change is lost as soon as the R session ends, so you'll need to re-run
it every time you start a new session (e.g. by adding it to a project
`.Rprofile`).

## Verify the installation

```{r,eval=FALSE}
Sys.which("diamond")
Sys.which("blastp")
```

Both should now return the full path to the executable instead of `""`.
You can additionally confirm that `orthologr` itself detects both tools:

```{r,eval=FALSE}
is_installed_diamond()
is_installed_blast()
```

```
[1] TRUE
[1] TRUE
```

## Additional information

`orthologr`'s interface functions (`diamond()`, `deepclust()`,
`deepclust_realign()`, `blast()`, `blast_best()`, `blast_rec()`, and their
underlying `is_installed_*()` checks) accept a `path` argument for cases
where you don't want to rely on `PATH` at all. Internally these functions
now call `normalizePath()` on any supplied `path`, so relative paths (e.g.
`path = "../installations"`) resolve correctly on Windows — this previously
failed because Windows' `system()` could not resolve relative executable
paths.

This fix applies to both `DIAMOND2` and `BLAST+`: `is_installed_blast()`
(used internally by `blast()`, `blast_best()`, and `blast_rec()`) now uses
the same `file.path()`-based approach, so passing a `path` argument works
correctly on Windows. If you prefer not to pass `path` every call, you can
make `blastp` permanently discoverable by adding the `BLAST+` `bin` folder
to the persistent Windows `PATH`:

1. Get the absolute path to the folder containing `blastp.exe`, e.g.
   `C:\Program Files\NCBI\blast-2.17.0+\bin`.
2. Open **Start Menu → "Edit the system environment variables"**.
3. Click **Environment Variables...**.
4. Under **User variables** (or **System variables** for all users), select
   `Path` and click **Edit...**.
5. Click **New** and paste the absolute folder path (no trailing backslash,
   no filename — just the directory).
6. Click **OK** on all dialogs to save, then **restart RStudio**
   (environment variable changes aren't picked up by already-running
   processes).
7. Verify with `Sys.which("blastp")`.

The same persistent approach also works for `DIAMOND2`, and is preferable
to the temporary session-only option if you use `orthologr` regularly.
